Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-04-05 paper.Rnw
Lots of updates: a2a_tracc results updated, mucho updates to analysis

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
#opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
#    fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw())
# bland_altman_plot <- function(dat){
#   dat$diff <- dat$pre - dat$post;
#   
#   return ggplot(dat, aes(pre, diff)) + 
#     geom_hline(yintercept = mean(dat$diff) + c(-2, 0, 2) * sd(dat$diff), 
#                linetype=2, color='brown') + 
#     geom_point() 
# }
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In previous work \citep{Chakravarty2012}, we applied \mb to segmentation
of the human striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. The contribution of the present work is to extend
our approach to the human hippocampus and perform a series of experiments to
rigorous validate the method. First, we conduct an extensive cross-validation of
\mb and basic multi-atlas segmentation on a subset of the ADNI1 dataset to
assess the accuracy of \mb under various parameter settings (number of
atlases and templates, registration and fusion methods). With the best
performing parameter configuration discovered above, we estimate \mb
intra-rater reliability by segmenting separately acquired T1 images of the
atlas subjects. For this experiment, we use the Winterburn atlases: digital
hippocampal subfield segmentations of five in-vivo high-resolution (300u
isotropic) T1-weighted MR scans\citep{Winterburn2013}. To validate \mb in a
real world situation, we segment the entire ADNI1 screening dataset and compare
our segmentations to established automated and manual segmentations.
Additionally, to ensure \mb accuracy across disease categories, we also
compare \mb segmentations to manual segmentations of 139 first episode
schizophrenic patients.

@@

\subsubsection{ADNI1 1.5T Screening Dataset}
Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI1 database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "ADNI1:Screening 1.5T"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008a}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a 192 x 192 x 166 matrix (x, y, and z directions) yielding a voxel
resolution of 1.25 x 1.25 x 1.2 $mm^3$. Clinical and demographic data are shown
in table \ref{tab:ADNI1_1.5T_Screening_Dataset_Demographics}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Screening Dataset Demographics Table                                  %
%                                                                             %
<<ADNI1-scr-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID < 2000,   
               method = "reverse", test=FALSE, overall=TRUE)
latex(tab, file="", size="scriptsize", 
    caption="ADNI1 1.5T Screening dataset demographics",
    title="tab:ADN1_1.5T_Screening_Dataset_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The hippocampus were traced following a validated protocol developed by Dr Jens
Pruessner (Pruessner et al., 2000). A recent update to this protocol by Dr J
Pruessner in 2006 allows to accurately and consistently subdivide the
hippocampus into three different subregions: head, body, and tail.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% First Episode SZ Demographics Table                                         %
%                                                                             %
<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", size="scriptsize", ctable=FALSE,
      caption="Schizophrenia First Episode Patient Demographics",
      title="tab:SZFEP_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

{\bf Dataset evaluated.} 69 1.5T images were arbitrarily selected from the {\em
ADNI1:Screening 1.5T} standardized dataset. 23 subjects were chosen from each
disease category (cognitively normal (CN), MCI and AD). Demographics for this
subset are shown in Table \ref{tab:ADNI1_xvalidation_demographics}.

@@

%%%%%
%%%%%               Experiments: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Screening Validation}
To test the accuracy of \mb on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

{\bf Dataset evaluated.} All images from the {\em ADNI1:Screening 1.5T}
standardized dataset.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Screening Template Library Demographics Table
%
<<ADNI1-scr-tmpllib-demographics, echo=F, results="asis",cache=FALSE>>=
scr_tmpllib_RIDs = c(0295, 0413, 0619, 0685, 0729, 0782, 0938, 0954, 1018, 1155,
                 0907, 0981, 0448, 0553, 0814, 1130, 0698, 1339, 1095, 0991,
                 0159)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               scr_tmpllib_RIDs, method = "reverse", test=FALSE, overall=TRUE)

latex(tab, landscape=FALSE, ctable=FALSE, file="", size="scriptsize",
      caption="ADNI1 Screening Template Library demographics", 
      title="tab:ADNI1_scr_tmpllib_demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%                                                                              %
% ADNI1 Screening Validation - results preparation                             %
%                                                                              %
<<ADN1-xval-prep, include=FALSE, cache=TRUE>>=
# all data from the validation trials
a_data       = read.csv(gzfile('data/a2a_ants/results_2013_01_03.csv.gz'))
#t_data       = read.csv(gzfile('data/a2a_tracc/xfmtrials_results_2013-04-01.csv.gz'))
t_data       = read.csv(gzfile('data/a2a_tracc/results_2012_12_31.csv.gz'))
diagnoses    = read.csv('data/a2a_diagnoses.csv')

# simplifies the table
t_data       = subset(t_data, atlases %in% c(1,3,5,7,9))
a_data       = subset(a_data, atlases %in% c(1,3,5,7,9))
t_data$se    = NULL
t_data$sn    = NULL
t_data$j     = NULL
a_data$se    = NULL
a_data$sn    = NULL
a_data$j     = NULL

@@

# separate MA majvote from MB so we can compute differences
# we only compare multiatlas using majority vote
t_ma = subset(t_data, approach=="multiatlas" & method=="majvote")
a_ma = subset(a_data, approach=="multiatlas" & method=="majvote")
t_mb = subset(t_data, approach=="mb")
a_mb = subset(a_data, approach=="mb")

@@

In this experiment we conducted 10 rounds of \mb and multi-atlas segmentation of
each of 69 subjects at a range of atlas and template library sizes, registration
algorithm (\ants or \animal), and three label fusion techniques.  Hippocampal
\mb-based segmentations using both \animal and \ants registration algorithm
demonstrate good overlap with manually derived gold-standards (Figure
\ref{fig:ADNI1-xval-k-mean}). Qualitatively, both \animal and \ants-based
segmentations demonstrate trend overlap accuracy that increases with the size of
atlas library and template library. With 12 or more templates \mb library of at
least 12 templates

No marked difference in segmentation accuracy is seen when either \animal or \ants
registration is used with different label fusion techniques, at any atlas or
template library sizes.  In every parameter configuration, the use of \mb with
\ants registration shows increased segmentation accuracy over \mb with \animal
registration. In the remainder of this section, only results using the \ants
registration algorithm will be shown.

@@

ggplot(all_data_mean, 
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1)) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" )
@
  \label{fig:ADNI1-xval-k_mean}
  \caption{
  {\em \mb mean DSC by atlas and template library size, registration algorithm,
  and label fusion method. Mean DSC is calculated over each subject from the
  ADNI1 Cross-Validation } 
  }
\end{figure}

@@

\begin{figure}
<<ADN1-Xval-k-diff, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ggplot(subset(all_data_mean, reg_method=="ANTS" & method.mb=="Majority Vote" &
                templates.mb %in% seq(1,20,2)), 
       aes(x=templates.mb, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_summary(fun.y=mean, geom="line", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  coord_cartesian(ylim=c(-0.1,0.05)) + 
  scale_x_continuous(breaks=seq(1,20,2)) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(legend=FALSE) + 
  scale_linetype_discrete(legend=FALSE) + 
  xlab( "Number of Templates" ) + 
  ylab( "Increase in mean Kappa between MAGeT brain and multiatlas" ) 
@
  \caption{Increase in mean Kappa between \mb and multi-atlas}
  \label{ADNI1-xval-k-diff}
\end{figure}

In addition to an improvement in accuracy over multi-atlas-based segmentation,
\mb also shows a decrease in the variability of segmentation accuracy
with different input atlases (Figure \todo{ref}).  Additionally, variability
decreases with the size of template library, and is significantly decreased
(p<0.05) when using 3-7 atlases and 17 or more templates.

\begin{figure}
<<ADN1-Xval-k-sd-t-test-p-plot, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ANTSmajvote = subset(all_data, reg_method=="ANTS" & method.mb=="Majority Vote")

@@

stats.tests = ddply(data.stats, c("atlases","templates.mb"), function (df) {
  t.var = t.test(df$MA.var, df$MB.var)
  t.sd  = t.test(df$MA.sd , df$MB.sd)
  data.frame(
    stat      = c("var", "sd"),
    statistic = c(t.var$statistic, t.sd$statistic),
    p.value   = c(t.var$p.value, t.sd$p.value)
  )
})

ggplot(stats.tests, 
       aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  facet_grid( . ~ stat) + 
  geom_line() + 
  geom_hline(aes(yintercept=0.05), linetype='dashed') + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + scale_y_continuous(breaks=seq(0,1,by=0.05)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" )

@
\end{figure}

<<ADNI1-Xval-sd-by-subject, cache=TRUE>>=
SDs = data.stats[c('subject','label','atlases','templates.mb','MA.sd','MB.sd')]
SDs = subset(SDs, templates.mb %in% c(5,10,15,20) & atlases == 5)

melted = melt(SDs, measure.vars=c('MA.sd','MB.sd'))
ggplot(melted, aes(x=templates.mb, y=value, 
                   group=interaction(templates.mb, variable), colour=as.factor(variable))) + 
  scale_x_continuous(breaks=c(5,10,15,20)) + 
  geom_violin() 
@

%- show variance in MAGeT performance by showing nine plots, each with mean multi-atlas performance plotted.

\begin{figure}
<<ADN1-Xval-k-diff-ANTS-20tmpls, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
a_20_tmpls = subset(all_data_mean, 
                      reg_method  == "ANTS" &
                      method.mb   == "Majority Vote" &
                      templates.mb== 20)
ggplot(a_20_tmpls, aes(x=factor(atlases))) +
  #geom_point(aes(y=k.mb,group=1,colour="MAGeT"),position="jitter") + 
  #geom_point(aes(y=k.ma,group=2,colour="MA"),position="jitter") + 
  geom_violin(aes(y=k.mb,group=factor(atlases),colour="MAGeT")) + 
  geom_violin(aes(y=k.ma,group=factor(atlases),colour="MA")) + 
  geom_smooth(aes(y=k.mb,group=1,colour="MAGeT")) +
  geom_smooth(aes(y=k.ma,group=2,colour="MA")) +

  #stat_summary(fun.y=mean,geom='point',size=3,colour="red",aes(y=k.ma)) +
  #stat_summary(fun.y=mean,geom='line',colour="red",aes(y=k.ma,group=1)) +
  #stat_summary(fun.y=mean,geom='point',size=3,colour="blue",aes(y=k.mb)) +
  #stat_summary(fun.y=mean,geom='line',colour="blue",aes(y=k.mb,group=1)) +
  labs(x="Number of Atlases", y = "Kappa", 
       title="mean Kappa of MA and MAGeT brain (ANTS/20 templates) \nvs\natlas library size") +
  scale_colour_hue(name="Method") + 
  coord_cartesian(ylim=c(0.7,0.9))
  #theme(legend.justification=c(1,0), legend.position=c(1,0))
@
  \caption{Change in mean Kappa between multi-atlas and \mb using \ants,
  20 templates}
  \label{}
\end{figure}

\begin{figure}
<<ADN1-Xval-ANTS-20tmpls-t-test-sd, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
a_20_tmpls = subset(all_data, 
                      reg_method  == "ANTS" &
                      method.mb    == "Majority Vote" &
                      templates.mb == 20 &
                      atlases     == 9 & 
                      volume.ma < 10000)
stddev = ddply(a_20_tmpls, c("subject", "label"), function (df) {
  data.frame(
    MA = sd(df$k.ma),  #MA
    MB = sd(df$k.mb)  #MB
  )
})

# TODO: against SNT
melted = melt(stddev, id.vars=c("subject", "label"))
ggplot(melted, aes(x=variable, y=value)) + geom_boxplot()
t.test(stddev$MA, stddev$MB)
@
  \caption{}
  \label{}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
#opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
#    fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw())

bland_altman_plot <- function(x, y) {
  data = data.frame(x=x,y=y,diff=x-y)
  return(ggplot(data, aes(x= x, y = diff)) + 
    geom_hline(yintercept = mean(data$diff) + c(-2, 0, 2) * sd(data$diff), 
               linetype=2, color='brown') + 
    geom_point())
}
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In previous work \citep{Chakravarty2012}, we applied \mb to segmentation
of the human striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. The contribution of the present work is to extend
our approach to the human hippocampus and perform a series of experiments to
rigorous validate the method. First, we conduct an extensive cross-validation of
\mb and basic multi-atlas segmentation on a subset of the ADNI1 dataset to
assess the accuracy of \mb under various parameter settings (number of
atlases and templates, registration and fusion methods). With the best
performing parameter configuration discovered above, we estimate \mb
intra-rater reliability by segmenting separately acquired T1 images of the
atlas subjects. For this experiment, we use the Winterburn atlases: digital
hippocampal subfield segmentations of five in-vivo high-resolution (300u
isotropic) T1-weighted MR scans\citep{Winterburn2013}. To validate \mb in a
real world situation, we segment the entire ADNI1 Complete 1Yr dataset and compare
our segmentations to established automated and manual segmentations.
Additionally, to ensure \mb accuracy across disease categories, we also
compare \mb segmentations to manual segmentations of 139 first episode
schizophrenic patients.

@@

\subsubsection{ADNI1 1.5T Complete 1Yr Dataset}
Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI1 database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "ADNI1:Complete 1Yr 1.5T"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008a}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a 192 x 192 x 166 matrix (x, y, and z directions) yielding a voxel
resolution of 1.25 x 1.25 x 1.2 $mm^3$. Clinical and demographic data are shown
in table \ref{tab:ADNI1_1.5T_Complete 1Yr_Dataset_Demographics}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Complete 1Yr Dataset Demographics Table                                  %
%                                                                             %
<<ADNI1-scr-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, 
               subset =  RID %in% yr1$RID & VISCODE %in% c('bl','m06','m12'),   
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize", 
    caption="ADNI1 1.5T_Complete_1Yr dataset demographics",
    title="tab:ADN1_1.5T_Complete_1Yr_Dataset_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The hippocampus were traced following a validated protocol developed by Dr Jens
Pruessner (Pruessner et al., 2000). A recent update to this protocol by Dr J
Pruessner in 2006 allows to accurately and consistently subdivide the
hippocampus into three different subregions: head, body, and tail.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% First Episode SZ Demographics Table                                         %
%

\begin{figure}
<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", size="scriptsize", ctable=FALSE,
      caption="Schizophrenia First Episode Patient Demographics",
      title="tab:SZFEP_Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \caption{
  }
\end{figure}

@@

{\bf Dataset evaluated.} 69 1.5T images were arbitrarily selected from the {\em
ADNI1:Complete 1Yr 1.5T} standardized dataset. 23 subjects were chosen from each
disease category (cognitively normal (CN), MCI and AD). Demographics for this
subset are shown in Table \ref{tab:ADNI1_xvalidation_demographics}.

@@

%%%%%
%%%%%               Experiments: ADNI-1 Complete 1Yr Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Complete 1Yr Validation}
To test the accuracy of \mb on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

{\bf Dataset evaluated.} All images from the {\em ADNI1:Complete 1Yr 1.5T}
standardized dataset.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Complete 1Yr Template Library Demographics Table
%
<<ADNI1-scr-tmpllib-demographics, echo=F, results="asis",cache=FALSE>>=
scr_tmpllib_RIDs = c(0295, 0413, 0619, 0685, 0729, 0782, 0938, 0954, 1018, 1155,
                 0907, 0981, 0448, 0553, 0814, 1130, 0698, 1339, 1095, 0991,
                 0159)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               scr_tmpllib_RIDs, method = "reverse", test=FALSE, overall=TRUE)

latex(tab, landscape=FALSE, ctable=FALSE, file="", size="scriptsize",
      caption="ADNI1 Complete 1Yr Template Library demographics", 
      title="tab:ADNI1_scr_tmpllib_demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%                                                                              %
% ADNI1 Complete 1Yr Validation - results preparation                             %
%                                                                              %
<<ADN1-xval-load, include=FALSE, cache=TRUE>>=
# all data from the validation trials
a_data       = read.csv(gzfile('data/a2a_ants/results_2013_01_03.csv.gz'))
#t_data       = read.csv(gzfile('data/a2a_tracc/xfmtrials_results_2013-04-01.csv.gz'))
t_data       = read.csv(gzfile('data/a2a_tracc/results_xfmjoin_2013_04_04.csv.gz'))
diagnoses    = read.csv('data/a2a_diagnoses.csv')
@

<<ADN1-xval-prep, include=FALSE, cache=TRUE>>=
# simplify the data
t_data       = subset(t_data, atlases %in% c(1,3,5,7,9))
a_data       = subset(a_data, atlases %in% c(1,3,5,7,9))
t_data$se    = NULL
t_data$sn    = NULL
t_data$j     = NULL
a_data$se    = NULL
a_data$sn    = NULL
a_data$j     = NULL

@@

# separate MA majvote from MB so we can compute differences
# we only compare multiatlas using majority vote
t_ma = subset(t_data, approach=="ma" & method=="majvote")
t_ma$approach="multiatlas"
a_ma = subset(a_data, approach=="multiatlas" & method=="majvote")
t_mb = subset(t_data, approach=="mb")
a_mb = subset(a_data, approach=="mb")

@@

In this experiment we conducted 10 rounds of \mb and multi-atlas segmentation of
each of 69 subjects at a range of atlas and template library sizes, registration
algorithm (\ants or \animal), and three label fusion techniques.  Hippocampal
\mb-based segmentations using both \animal and \ants registration algorithm
demonstrate good overlap with manually derived gold-standards (Figure
\ref{fig:ADNI1-xval-k-mean}). Qualitatively, both \animal and \ants-based
segmentations demonstrate trend overlap accuracy that increases with the size of
atlas library and template library. Improvement in accuracy diminishes noticeably
with template libraries larger than ten images.

No marked difference in segmentation accuracy is seen when either \animal or \ants
registration is used with different label fusion techniques, at any atlas or
template library sizes.  In every parameter configuration, the use of \mb with
\ants registration shows a pronounced increase in segmentation accuracy over \mb with \animal
registration. In the remainder of this section, only results using the \ants
registration algorithm will be shown.

@@

ggplot(all_data_mean, 
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1)) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" )
@
  \label{fig:ADNI1-xval-k_mean}
  \caption{
  {\em Mean DSC by atlas and template library size, registration algorithm,
  and label fusion method. The mean DSC score is calculated over all subjects 
  from the ADNI1 Cross-Validation dataset for each parameter setting.} 
  }
\end{figure}

@@

\begin{figure}
<<ADN1-Xval-k-diff, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ggplot(subset(all_data_mean, reg_method=="ANTS" & method.mb=="Majority Vote" &
                templates.mb %in% seq(1,20,2)), 
       aes(x=templates.mb, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_summary(fun.y=mean, geom="line", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  coord_cartesian(ylim=c(-0.1,0.05)) + 
  scale_x_continuous(breaks=seq(1,20,2)) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(legend=FALSE) + 
  scale_linetype_discrete(legend=FALSE) + 
  xlab( "Number of Templates" ) + 
  ylab( "Increase in mean DSC between MAGeT brain and multiatlas" ) 
@
  \caption{The difference in mean DSC between \mb and multi-atlas segmentations for 
  a range of parameter settings.}
  \label{ADNI1-xval-k-diff}
\end{figure}

In addition to an improvement in accuracy over multi-atlas-based segmentation,
\mb also shows a decrease in the variability of segmentation accuracy 
(Figure \todo{ADNI1-xval-variability}).  The size of template library 
necessary to reach a significant (p<0.5) decrease in variance and standard 
deviation grows with the size of atlas library used.  A template library of 19
images is sufficient to show significant decrease in variance and standard 
deviation for 3-7 atlases.

\begin{figure}
<<ADN1-Xval-variability, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ANTSmajvote = subset(all_data, 
                     reg_method=="ANTS" & 
                     method.mb=="Majority Vote" &
                    templates.mb %in% seq(1,20,2))

@@

stats.tests = ddply(data.stats, c("atlases","templates.mb"), function (df) {
  t.var = t.test(df$MA.var, df$MB.var)
  t.sd  = t.test(df$MA.sd , df$MB.sd)
  data.frame(
    stat      = c("Variance", "Standard Deviation"),
    statistic = c(t.var$statistic, t.sd$statistic),
    p.value   = c(t.var$p.value, t.sd$p.value),
    direction = c(t.var$statistic < 0, t.sd$statistic < 0)
  )
})

ggplot(stats.tests, 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases), 
      size = abs(statistic))) + 
  facet_grid( . ~ stat) + 
  geom_line(aes(size=0.1)) + 
  geom_point(aes(shape = factor(direction))) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  #scale_size_continuous(guide='none') + 
  scale_shape_discrete("t-statistic sign", labels=c("positive","negative")) + 
  scale_size_area(name="t-statistic magnitude") + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) 
@
  \caption{Difference in variability of \mb vs. multi-atlas segmentation accuracy. 
  In these plots we consider the segmentation accuracy within each subject across
  the 10 rounds of validation per parameter setting. Standard deviation and 
  variance is computed per subject for \mb and multi-atlas, and the distribution
  of these test statistics is compared with a t-test. The p-value of this test, 
  is shown on the y-axis (scaled logarithmically).}
  \label{ADNI1-xval-variability}
\end{figure}

% <<ADNI1-Xval-sd-by-subject, cache=TRUE>>=
% SDs = data.stats[c('subject','label','atlases','templates.mb','MA.sd','MB.sd')]
% SDs = subset(SDs, templates.mb %in% c(5,10,15,20) & atlases == 5)
% 
% melted = melt(SDs, measure.vars=c('MA.sd','MB.sd'))
% ggplot(melted, aes(x=templates.mb, y=value, 
%                    group=interaction(templates.mb, variable), colour=as.factor(variable))) + 
%   scale_x_continuous(breaks=c(5,10,15,20)) + 
%   geom_violin() 
% @
% \begin{figure}
% <<ADN1-Xval-k-diff-ANTS-20tmpls, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
% a_20_tmpls = subset(all_data_mean, 
%                       reg_method  == "ANTS" &
%                       method.mb   == "Majority Vote" &
%                       templates.mb== 20)
% ggplot(a_20_tmpls, aes(x=factor(atlases))) +
%   #geom_point(aes(y=k.mb,group=1,colour="MAGeT"),position="jitter") + 
%   #geom_point(aes(y=k.ma,group=2,colour="MA"),position="jitter") + 
%   geom_violin(aes(y=k.mb,group=factor(atlases),colour="MAGeT")) + 
%   geom_violin(aes(y=k.ma,group=factor(atlases),colour="MA")) + 
%   geom_smooth(aes(y=k.mb,group=1,colour="MAGeT")) +
%   geom_smooth(aes(y=k.ma,group=2,colour="MA")) +
%   
%   #stat_summary(fun.y=mean,geom='point',size=3,colour="red",aes(y=k.ma)) +
%   #stat_summary(fun.y=mean,geom='line',colour="red",aes(y=k.ma,group=1)) +
%   #stat_summary(fun.y=mean,geom='point',size=3,colour="blue",aes(y=k.mb)) +
%   #stat_summary(fun.y=mean,geom='line',colour="blue",aes(y=k.mb,group=1)) +
%   labs(x="Number of Atlases", y = "Kappa", 
%        title="mean Kappa of MA and MAGeT brain (ANTS/20 templates) \nvs\natlas library size") +
%   scale_colour_hue(name="Method") + 
%   coord_cartesian(ylim=c(0.7,0.9))
%   #theme(legend.justification=c(1,0), legend.position=c(1,0))
% @
%   \caption{Change in mean Kappa between multi-atlas and \mb using \ants,
%   20 templates}
%   \label{}
% \end{figure}
% 
% \begin{figure}
% <<ADN1-Xval-ANTS-20tmpls-t-test-sd, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
% a_20_tmpls = subset(all_data, 
%                       reg_method  == "ANTS" &
%                       method.mb    == "Majority Vote" &
%                       templates.mb == 20 &
%                       atlases     == 9 & 
%                       volume.ma < 10000)
% stddev = ddply(a_20_tmpls, c("subject", "label"), function (df) {
%   data.frame(
%     MA = sd(df$k.ma),  #MA
%     MB = sd(df$k.mb)  #MB
%   )
% })
% 
% 
% # TODO: against SNT
% melted = melt(stddev, id.vars=c("subject", "label"))
% ggplot(melted, aes(x=variable, y=value)) + geom_boxplot()
% t.test(stddev$MA, stddev$MB)
% @
%   \caption{}
%   \label{}
% \end{figure}
% 
%
